Simple repression draft

### 2020-05-17 doc/paper.tex
Additions to text on constit promoter inference

Let $p_{m,R}(t)$ denote the joint probability of finding the
system in the repressor bound state with $m$ mRNA molecules
present at time $t$, and similarly define $p_{m,U}(t)$ for the
repressor unbound state. 
This model is governed by a master equation giving the time
evolution of $p_{m,R}(t)$ and $p_{m,U}(t)$, which is
\begin{align}
\begin{split}
\deriv{t}p_{m,R}(t) =& - k_R^- p_{m,R}(t) + k_R^+ p_{m,U}(t)
                + (m+1)\gamma p_{m+1,R}(t) - \gamma p_{m,R}(t)
\\
\deriv{t}p_{m,U}(t) =&\; k_R^- p_{m,R}(t) - k_R^+ p_{m,U}(t)
                        + rp_{m-1,U}(t) - rp_{m,U}(t)
                \\
                &+ (m+1)\gamma p_{m+1,U}(t) - \gamma p_{m,U}(t).
\end{split}
\end{align}
In each equation, the first two terms describe transitions
between promoter states due to repressors unbinding and binding,
respectively. The final two terms describe degradation of mRNA,
decreasing the copy number by one, and the terms with coefficient
$r$ describe transcription initiation increasing the mRNA copy
number by one.

We can greatly simplify the notation, which will be especially
useful for the more complicated models yet to come, by
re-expressing the master equation in vector form. The promoter
states are collected into a vector and the rate constants are
collected into matrices as
\begin{equation}
\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,U} \end{pmatrix},\
\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ \\ k_R^- & -k_R^+ \end{pmatrix},\
\matr{R} = \begin{pmatrix} 0 & 0 \\ 0 & r \end{pmatrix},\
\label{eq:2state_cme_matrices}
\end{equation}
so that the master equation may be condensed as
\begin{equation}
\deriv{t}\vect{p}_m(t) =
\left( \matr{K} - \matr{R} - \gamma m \matr{I} \right) \vect{p}_m(t)
                + \matr{R} \vect{p}_{m-1}(t)
                + \gamma (m+1) \matr{I} \vect{p}_{m+1}(t)
                - \gamma m \matr{I} \vect{p}_m(t),
\label{eq:2state_rep_cme}
\end{equation}
where $\matr{I}$ is the identity matrix.
Taking steady state by setting time derivatives to zero,
the mean mRNA can be found to be
\begin{equation}
\langle m \rangle = \frac{r}{\gamma}
        \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
\end{equation}
with the algebra details again deferred to the appendix.
Recall $k_R^+$ is proportional to the repressor copy number, so
in computing fold-change, absence of repressor corresponds to
$k_R^+\rightarrow0$.
Therefore fold-change in this model is simply
\begin{equation}
\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
\end{equation}
again matching the master curve of~\fig{fig1:means_cartoons} with $\rho=1$.

@@

The master equation governing this model is analogous
to~\eqrange{eq:2state_cme_matrices}{eq:2state_rep_cme} for model 1 above.
The main subtlety arises since transcription initiation
accompanies a promoter state change.
This can be understood by analogy to $\matr{K}$. The
off-diagonals and diagonals of $\matr{K}$ correspond to
transitions arriving at or departing from, respectively, the
promoter state of interest. If transcription initiations are
accompanied by promoter state changes, we must have separate
matrices for arriving and departing transcription events since
the arriving and departing transitions have different initial
copy number of mRNA, unlike for $\matr{K}$ where they are the same.
The master equation for this model is
\begin{equation}
\deriv{t}\vect{p}_m(t) =
\left( \matr{K} - \matr{R_D} - \gamma m \matr{I} \right) \vect{p}_m(t)
                + \matr{R_A} \vect{p}_{m-1}(t) +
                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t)
                - \gamma m \matr{I} \vect{p}_m(t),
\label{eq:3state_rep_cme}
\end{equation}
with the state vector and promoter transition matrix defined as
\begin{equation}
\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\ p_{m,P} \end{pmatrix},\
\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 \\
                        k_R^- & -k_R^+ -k_P^+ & k_P^- \\
                        0 & k_P^+ & -k_P^- 
                \end{pmatrix},
\label{eq:3state_cme_matrices_pt1}
\end{equation}
and the initiation matrices given by
\begin{equation}
\matr{R_A} = \begin{pmatrix}
                0 & 0 & 0 \\ 0 & 0 & r \\ 0 & 0 & 0
                \end{pmatrix},\
\matr{R_D} = \begin{pmatrix}
                0 & 0 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & r
                \end{pmatrix},
\label{eq:3state_cme_matrices_pt2}
\end{equation}
The elements of $\vect{p}_m$ encode the probabilities of having
$m$ mRNA present along with the promoter having repressor bound,
being empty, or having RNAP bound, respectively.
$\matr{R_A}$ describes probability flux \textit{arriving} at the state
$\vect{p}_m$ from a state with one fewer mRNA, namely
$\vect{p}_{m-1}$, and $\matr{R_D}$ describes probability flux
\textit{departing} from the state $\vect{p}_m$
for a state with one more mRNA, namely $\vect{p}_{m+1}$.
$\matr{K}$ is closely analogous to model 1.

@@

The master equation governing this model is
identical in form to model 2 above, namely
\begin{equation}
\deriv{t}\vect{p}_m(t) =
\left( \matr{K} - \matr{R_D} - \gamma m \matr{I} \right) \vect{p}_m(t)
                + \matr{R_A} \vect{p}_{m-1}(t) +
                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t)
                - \gamma m \matr{I} \vect{p}_m(t),
\end{equation}
but obviously with a higher-dimensional state space and different matrices.
The state vector and promoter transition matrix are now
\begin{equation}
\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\
                             p_{m,C} \\ p_{m,O}\end{pmatrix},\
\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 & 0\\
                        k_R^- & -k_R^+ -k_P^+ & k_P^- & 0 \\
                        0 & k_P^+ & -k_P^- - k_O & 0 \\
                        0 & 0 & k_O & 0
                \end{pmatrix},
\end{equation}
with the four promoter states, in order, being repressor bound,
empty, RNAP closed complex, and RNAP open complex.
Besides increasing dimension by one again, the only new feature in
$\matr{K}$ is the rate $k_O$, representing the (irreversible)
rate of open complex formation from the closed complex.
The initiation matrices are given by
\begin{equation}
\matr{R_A} = \begin{pmatrix}
        0 & 0 & 0 & 0 \\ 0 & 0 & 0 & r \\ 0 & 0 & 0 & 0\\ 0 & 0 & 0 & 0
                \end{pmatrix},\
\matr{R_D} = \begin{pmatrix}
        0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0\\ 0 & 0 & 0 & r
                \end{pmatrix},
\end{equation}
again closely analogous to model 2.

@@

\subsection{Parameter inference for constitutive promoters}
Having established that a one-state bursty model achieves a
Goldilocks level of complexity, we test it more thoroughly on all
the data from constitutive promoters from~\cite{Jones2014}.
\fig{fig:constit_post} shows the results.
\begin{figure}%[h!]
\centering
\includegraphics[width=0.9\textwidth]{../figures/fig2/fig2pt1.pdf}
\caption{\textbf{Constitutive promoter posterior
        inference and predictive checks.}
        (A) shows the joint posterior density from MCMC sampling of model 5,
        bursty promoter with negative binomially-distributed steady state.
        (B) shows the ECDF of the observed population distribution
        of mRNA transcripts under the control of a constitutive lacUV5 promoter.
        Posterior predictive percentiles are shown as colored bands for
        model (1), Poisson, and model (5), negative binomial.
        }
\label{fig:uv5_post}
\end{figure}

@@

\subsection{Transcription factor kinetics can be inferred from FISH measurements}
Figure~3 presents inferred transcription factor rates, compares
them to single-molecule measurements, and compares their ratios
to gold-standard binding energies.
\begin{figure}%[h!]
\centering
\includegraphics[width=0.9\textwidth]{../figures/fig3/fig3.ai}
\caption{\textbf{Simple repression model comparison.}
(A) depicts the cartoon of simple repression for which we infer parameters.
(B) shows 50\% and 95\% contours of several 2D slices of the 9D posterior
distribution, from fitting one unbinding rate for each operator (O1, O2, O3)
and one binding rate for each aTc concentration (corresponding to an
unknown mean repressor copy number).
(C) compares ratios of our inferred unbinding rates with operator
binding energy differences measured in Garcia \& Phillips 2011 (triangles)
and Razo-Mejia et.\ al.\ 2018 (squares).
(D) compares our inferred unbinding rates with single-molecule measurements
of the same from Hammar et.\ at.\ 2014 \mmnote{Johan Elf's group}.
Values are of similar order of magnitude,
but a precise quantitative comparison unfortunately depends sensitively
on the assumed mRNA lifetime, which was not measured.
    }
\label{fig3:kR_inferences}
\end{figure}

@@

>>>>

Let $p_{m,R}(t)$ denote the joint probability of finding the
system in the repressor bound state with $m$ mRNA molecules
present at time $t$, and similarly define $p_{m,U}(t)$ for the
repressor unbound state. 
This model is governed by a master equation giving the time
evolution of $p_{m,R}(t)$ and $p_{m,U}(t)$, which is
\begin{align}
\begin{split}
\deriv{t}p_{m,R}(t) =& - k_R^- p_{m,R}(t) + k_R^+ p_{m,U}(t)
                + (m+1)\gamma p_{m+1,R}(t) - \gamma p_{m,R}(t)
\\
\deriv{t}p_{m,U}(t) =&\; k_R^- p_{m,R}(t) - k_R^+ p_{m,U}(t)
                        + rp_{m-1,U}(t) - rp_{m,U}(t)
                \\
                &+ (m+1)\gamma p_{m+1,U}(t) - \gamma p_{m,U}(t).
\label{eq:poisson_promoter_cme}
\end{split}
\end{align}
In each equation, the first two terms describe transitions
between promoter states due to repressors unbinding and binding,
respectively. The final two terms describe degradation of mRNA,
decreasing the copy number by one, and the terms with coefficient
$r$ describe transcription initiation increasing the mRNA copy
number by one.

We can greatly simplify the notation, which will be especially
useful for the more complicated models yet to come, by
re-expressing the master equation in vector form. The promoter
states are collected into a vector and the rate constants are
collected into matrices as
\begin{equation}
\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,U} \end{pmatrix},\
\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ \\ k_R^- & -k_R^+ \end{pmatrix},\
\matr{R} = \begin{pmatrix} 0 & 0 \\ 0 & r \end{pmatrix},\
\label{eq:2state_cme_matrices}
\end{equation}
so that the master equation may be condensed as
\begin{equation}
\deriv{t}\vect{p}_m(t) =
\left( \matr{K} - \matr{R} - \gamma m \matr{I} \right) \vect{p}_m(t)
                + \matr{R} \vect{p}_{m-1}(t)
                + \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
\label{eq:2state_rep_cme}
\end{equation}
where $\matr{I}$ is the identity matrix.
Taking steady state by setting time derivatives to zero,
the mean mRNA can be found to be
\begin{equation}
\langle m \rangle = \frac{r}{\gamma}
        \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
\end{equation}
with the algebra details again deferred to the appendix.
Recall $k_R^+$ is proportional to the repressor copy number, so
in computing fold-change, absence of repressor corresponds to
$k_R^+\rightarrow0$.
Therefore fold-change in this model is simply
\begin{equation}
\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
\end{equation}
again matching the master curve of~\fig{fig1:means_cartoons} with $\rho=1$.

@@

The master equation governing this model is analogous
to~\eqrange{eq:2state_cme_matrices}{eq:2state_rep_cme} for model 1 above.
The main subtlety arises since transcription initiation
accompanies a promoter state change.
This can be understood by analogy to $\matr{K}$. The
off-diagonals and diagonals of $\matr{K}$ correspond to
transitions arriving at or departing from, respectively, the
promoter state of interest. If transcription initiations are
accompanied by promoter state changes, we must have separate
matrices for arriving and departing transcription events since
the arriving and departing transitions have different initial
copy number of mRNA, unlike for $\matr{K}$ where they are the same.
The master equation for this model is
\begin{equation}
\deriv{t}\vect{p}_m(t) =
\left( \matr{K} - \matr{R_D} - \gamma m \matr{I} \right) \vect{p}_m(t)
                + \matr{R_A} \vect{p}_{m-1}(t) +
                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
\label{eq:3state_rep_cme}
\end{equation}
with the state vector and promoter transition matrix defined as
\begin{equation}
\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\ p_{m,P} \end{pmatrix},\
\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 \\
                        k_R^- & -k_R^+ -k_P^+ & k_P^- \\
                        0 & k_P^+ & -k_P^- 
                \end{pmatrix},
\label{eq:3state_cme_matrices_pt1}
\end{equation}
and the initiation matrices given by
\begin{equation}
\matr{R_A} = \begin{pmatrix}
                0 & 0 & 0 \\ 0 & 0 & r \\ 0 & 0 & 0
                \end{pmatrix},\
\matr{R_D} = \begin{pmatrix}
                0 & 0 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & r
                \end{pmatrix},
\label{eq:3state_cme_matrices_pt2}
\end{equation}
The elements of $\vect{p}_m$ encode the probabilities of having
$m$ mRNA present along with the promoter having repressor bound,
being empty, or having RNAP bound, respectively.
$\matr{R_A}$ describes probability flux \textit{arriving} at the state
$\vect{p}_m$ from a state with one fewer mRNA, namely
$\vect{p}_{m-1}$, and $\matr{R_D}$ describes probability flux
\textit{departing} from the state $\vect{p}_m$
for a state with one more mRNA, namely $\vect{p}_{m+1}$.
$\matr{K}$ is closely analogous to model 1.

@@

The master equation governing this model is
identical in form to model 2 above, namely
\begin{equation}
\deriv{t}\vect{p}_m(t) =
\left( \matr{K} - \matr{R_D} - \gamma m \matr{I} \right) \vect{p}_m(t)
                + \matr{R_A} \vect{p}_{m-1}(t) +
                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
\end{equation}
but obviously with a higher-dimensional state space and different matrices.
The state vector and promoter transition matrix are now
\begin{equation}
\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\
                             p_{m,C} \\ p_{m,O}\end{pmatrix},\
\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 & 0\\
                        k_R^- & -k_R^+ -k_P^+ & k_P^- & 0 \\
                        0 & k_P^+ & -k_P^- - k_O & 0 \\
                        0 & 0 & k_O & 0
                \end{pmatrix},
\end{equation}
with the four promoter states, in order, being repressor bound,
empty, RNAP closed complex, and RNAP open complex.
Besides increasing dimension by one again, the only new feature in
$\matr{K}$ is the rate $k_O$, representing the (irreversible)
rate of open complex formation from the closed complex.
The initiation matrices are given by
\begin{equation}
\matr{R_A} = \begin{pmatrix}
        0 & 0 & 0 & 0 \\ 0 & 0 & 0 & r \\ 0 & 0 & 0 & 0\\ 0 & 0 & 0 & 0
                \end{pmatrix},\
\matr{R_D} = \begin{pmatrix}
        0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0\\ 0 & 0 & 0 & r
                \end{pmatrix},
\end{equation}
again closely analogous to model 2.

@@

\section{Finding the ``right'' model: Bayesian parameter inference and model selection}
\subsection{Parameter inference for constitutive promoters}

From consideration of Fano factors in the previous section, we
suspect that model 5 in~\fig{fig:constit_cartoons},
a one-state bursty model of constitutive promoters,
achieves a Goldilocks level of complexity.
Does this stand up to closer scrutiny, namely, comparison to
full mRNA distributions rather than simply their moments?
We will test this thoroughly on all the data
from constitutive promoters from Jones et.\ al.~\cite{Jones2014}.

It will be instructive, however, to first consider the
Poisson promoter, model 1 in~\fig{fig:constit_cartoons}.
This is a convenient proxy for models 2 and 3, whose exact
steady-state mRNA distributions would be tedious to calculate
and we will not explicitly consider.
But since they have Fano factors $\nu\le 1$, they can
fare no better than the simple Poisson model.
We will also not explicitly consider model 4 from~\fig{fig:constit_cartoons}
since it was already thoroughly analyzed in~\cite{Razo-Mejia2020},
and since model 5 can be viewed as a special case of it.

In performing parameter inference on the FISH data
of~\cite{Jones2014}, building a generative model is
straightforward. The data are single-cell mRNA counts, and we
assume different cells to be independent and identically
distributed from the underlying generative model, which is just
the steady-state mRNA distribution computed from the master
equation of the underlying model.

\subsubsection{Model 1: Poisson promoter}
\mmnote{A lot of the details I've written here might be excessive for main text and serve better as an intro-by-example to Bayes in the appendix, then reduce main text to a bit more bare bones?}
For this model the master equation of interest
is~\eq{eq:poisson_promoter_cme} with repressor set to zero, i.e.,
\begin{equation}
\deriv{t}p_{m,U}(t) = rp_{m-1,U}(t) - rp_{m,U}(t)
        + (m+1)\gamma p_{m+1,U}(t) - \gamma p_{m,U}(t),
\end{equation}
whose steady-state distribution is Poisson, given by
\begin{equation}
p_{m,U} = \frac{\lambda^m e^{-\lambda}}{m!}
\end{equation}
with $\lambda=r/\gamma$. This steady state distribution defines our
likelihood $p(m\mid\lambda)$, i.e.,
\begin{equation}
p(m\mid\lambda) = \frac{\lambda^m e^{-\lambda}}{m!}
\label{eq:poisson_inference010}
\end{equation}
is the probability of observing
a single cell with $m$ mRNAs given a known value of $\lambda$.

With our likelihood in hand, we can invoke Bayes theorem to write down
the posterior distribution
on our model parameter $\lambda$, given by
\begin{equation}
p(\lambda\mid D) \propto p(D\mid\lambda) p(\lambda),
\end{equation}
where our data $D=\{m_1, m_2,\dots, m_N\}$ is a list of mRNA counts
over a population of $N$ identical cells.
As is standard we have neglected the factor $p(D)$ on the right
hand side since it is independent of $\lambda$ and serves only as
a normalization factor.
Since we assume each cell's mRNA count is independent of others,
the likelihood is simply a product of the
single cell likelihoods given by \eq{eq:poisson_inference010} above, so
\begin{equation}
p(D\mid\lambda) = \prod_{k=1}^N \frac{\lambda^{m_k}e^{-\lambda}}{m_k!}.
\end{equation}
This is often notated simply as
\begin{equation}
D\mid\lambda \sim \text{Poisson}(\lambda),
\end{equation}
which is read as the data $D$, conditioned on a value of
$\lambda$, is Poisson distributed with mean $\lambda$.
The $\sim$ operator literally denotes ``is distributed according to.''

To proceed we need to specify a prior. In this case we are extremely data-rich,
as the dataset from Jones et.\ al~\cite{Jones2014} has of order
1000-3000 single-cell measurements for each promoter,
so our choice of prior matters little here,
as long as it does not exclude the true value of $\lambda$.
A convenient choice for our Poisson likelihood is its conjugate prior,
the gamma distribution. Although they only exist for
certain special likelihoods, conjugate priors have the nice
feature that the posterior takes the same functional form as the
prior, with updated parameters. This makes calculations
analytically tractable and also offers a nice interpretation of
the inference procedure as updating our knowledge about the model
parameters.

Putting a gamma prior on $\lambda$ introduces two new parameters
$\alpha$ and $\beta$ which parametrize the gamma distribution
itself, which we use to encode the range of $\lambda$ values we
view as reasonable. Recall $\lambda$ is the mean steady-state
mRNA count per cell, which \textit{a priori} could plausibly be
anywhere from 0 to a few hundred. $\alpha=1$ and $\beta=1/50$
achieve this, since the gamma distribution is strictly positive
with mean $\alpha/\beta$ and standard deviation
$\sqrt{\alpha}/\beta$. To be explicit, then, our prior is
\begin{equation}
\lambda\mid\alpha,\beta \sim \text{Gamma}(\alpha, \beta)
\end{equation}

As an aside, note that if we did not know that our prior was a
conjugate prior, we could still write down our posterior
distribution from its definition as
\begin{equation}
p(\lambda\mid D,\alpha,\beta)
\propto p(D\mid\lambda) p(\lambda \mid\alpha,\beta)
\propto \left(\prod_{k=1}^N \frac{\lambda^{m_k}e^{-\lambda}}{m_k!}\right)
        \frac{\beta}{\Gamma(\alpha)}(\beta\lambda)^{\alpha-1} e^{-\beta\lambda}
.
\end{equation}
Without foreknowledge that this in fact reduces to a gamma distribution,
this expression might appear rather inscrutable. When conjugate
priors are unavailable for the likelihood of interest - which is
almost always the case for models with $>1$ model parameter -
this inscrutability is the norm, and making sense of posteriors
analytically is almost always impossible. Fortunately, MCMC
sampling provides us a powerful method of constructing posteriors
numerically which we will make use of extensively.

Since we did use a conjugate prior, we may simply look up our
posterior in any standard table of conjugate priors-likelihood
pairs, from which we find that
\begin{equation}
\lambda\mid D,\alpha,\beta
\sim \text{Gamma}\left(\alpha + \bar{m}N, \beta + N\right),
\end{equation}
where we defined the sample mean $\bar{m} = \frac{1}{N}\sum_k m_k$
for notational convenience.
A glance at the FISH data from~\cite{Jones2014} reveals that $N$ is
$\mathcal{O}(10^3)$ and $\langle m\rangle \gtrsim 0.1$ for all
constitutive strains in~\cite{Jones2014}, so
$\bar{m}N \gtrsim 10^2$. Therefore as we suspected,
our prior parameters are completely overwhelmed by the data.
In fact, $\bar{m}N$ and $N$ are so large that we can,
to an excellent approximation, ignore the $\alpha$ and $\beta$ dependence and
approximate the gamma distribution as a Gaussian with mean $\bar{m}$ and
standard deviation $\sqrt{\bar{m}/N}$, giving
\begin{equation}
\lambda\mid D
\sim \text{Gamma}\left(\alpha + \bar{m}N, \beta + N\right)
\approx \text{Normal}\left(\bar{m}, \sqrt{\frac{\bar{m}}{N}}\right).
\end{equation}
As an example with real numbers, for the \textit{lacUV5} promoter,
Jones et.\ al~\cite{Jones2014} measured 2648 cells with an average
mRNA count per cell of $\bar{m} \approx 18.7$. In this case then,
our posterior is
\begin{equation}
\lambda\mid D_{UV5}
\sim \text{Normal}\left(18.7, 0.08\right),
\end{equation}
which suggests we have inferred our model's one parameter to a precision of order 1\%.

This is not wrong, but it is not the full story. The model's posterior is tightly constrained, but is it a good generative model? In other words, does the model generate data that ``look like'' our actual data, and is it therefore plausible that the model captures the important features of the data generating process? This intuitive notion can be codified with \textit{posterior predictive checks}, or PPCs, and we will see that this simple Poisson model fails badly.
\begin{figure}%[h!]
\centering
\includegraphics[width=0.9\textwidth]{../figures/fig2/fig2pt1.pdf}
\caption{\textbf{Constitutive promoter posterior
        inference and predictive checks.}
        (A) shows the joint posterior density from MCMC sampling of model 5,
        bursty promoter with negative binomially-distributed steady state.
        (B) shows the ECDF of the observed population distribution
        of mRNA transcripts under the control of a constitutive lacUV5 promoter.
        Posterior predictive percentiles are shown as colored bands for
        model (1), Poisson, and model (5), negative binomial.
        }
\label{fig:uv5_post}
\end{figure}

\mmnote{Outline:
\begin{itemize}
\item Poisson model. Write Bayes, briefly explain conjugacy and it's
convenience. immediately write down posterior, make Gaussian
approx. Explain PPC sampling (actually PP dist is analytically
known but you still have to draw RNGs from it. Also this obscures
the intuitive meaning of PPC sampling from posterior and
generating synthetic data, so I think it's more useful to just
generate PP samples from posterior draws, this gives a more
intuitive apples-to-apples comparison w/ model 5).
\item Bursty model. Write Bayes, sketch story of NB likelihood, sketch choice
of priors, show~\fig{fig:uv5_post}, see bursty $\gg$ Poisson model.
\item All 18 promoters. Hey the model still works on things other than
UV5, cool. Interesting and somewhat counterintuitive scaling b/w
burst rate and binding energy. Puzzle in comparing w/ Chong2014
supercoiling model. Are we missing important things?? Unclear, we
leave as open question. Good enough for our purposes: posterior is identifiable, PPC is great, and of the models we've thought of it is unique in satisfying both.
\end{itemize}
}

@@

\subsection{Transcription factor kinetics can be inferred from FISH measurements}
\mmnote{Outline:
\begin{itemize}
\item Full model.
Write Bayes, handwave 2F1 dist story in limits of weak, strong,
and intermediate repression.
\item Hey look it mostly works.
With full model, 9D model is well identified, though most of the
individual experiments unsurprisingly are not: only the ratio of
repressor rates is identifiable. PPC is not perfect, it kinda
misses some rep/op pairs but overall it's surprisingly predictive
considering how strong were the assumptions that went into it.
\item Do we chalk this up to experimental imperfections or is there
interesting biology to be found in the disagreements?
I'm inclined towards the former. Even just catching cells in
truly apples-to-apples growth phases, or matching up parameters
of imaging sessions weeks or months apart, is hardly trivial.
\item The alternative is to refine our model and/or add on
complexity, and it's not at all clear what additions to make to
the model, especially how to produce the long tail of repressed
strains over UV5. I see no way to get that without getting into
complicated time-correlation/history effects, where
transcription...rebounds?? After being repressed?? Sounds odd.
\item Comparison between my inferred rates, equilibrium binding E, and single-molecule measurements. Not quite as slam-dunk as I'd hoped because of $\gamma$ dependence, but rates are w/in factor of 2 or better, seems quite good really! With exception of one Oid data data point, free energies are w/in about 0.5 kT also. That's not massively larger than the repeatability threshold of fold-change measurements, at maybe 0.1 to 0.3 kT.
\item Surprising that so many model parameters can be inferred from just a few distributions that, by eye, don't have strong features! Does require some assumptions, e.g., about equivalence of rates across experiments, but that seems quite reasonable.
\end{itemize}
}

Figure~3 presents inferred transcription factor rates, compares
them to single-molecule measurements, and compares their ratios
to gold-standard binding energies.
\begin{figure}%[h!]
\centering
\includegraphics[width=0.9\textwidth]{../figures/fig3/fig3.ai}
\caption{\textbf{Simple repression model comparison.}
(A) depicts the cartoon of simple repression for which we infer parameters.
\mmnote{
Not sure panel (A) is necessary, we've already shown it in at least
2 other figs. We could instead show some alternative slices of 9D posterior
as a companion to (B), or something totally different\dots??}
(B) shows 50\% and 95\% contours of several 2D slices of the 9D posterior
distribution, from fitting one unbinding rate for each operator (O1, O2, O3)
and one binding rate for each aTc concentration (corresponding to an
unknown mean repressor copy number).
(C) compares ratios of our inferred unbinding rates with operator
binding energy differences measured in Garcia \& Phillips 2011 (triangles)
and Razo-Mejia et.\ al.\ 2018 (squares).
(D) compares our inferred unbinding rates with single-molecule measurements
of the same from Hammar et.\ at.\ 2014 \mmnote{Johan Elf's group}.
Values are of similar order of magnitude,
but a precise quantitative comparison unfortunately depends sensitively
on the assumed mRNA lifetime, which was not measured.
    }
\label{fig3:kR_inferences}
\end{figure}

@@

\section{Bayesian inference}
\mmnote{
\begin{itemize}
\item Bayes theorem.
Elaborate on posterior as plausibility over
space of model parameters.
\item Choosing likelihoods.
Easy here, just the steady-state distributions of our master
equations, assuming iid single-cell measurements.
\item Choosing priors.
Not critical here because we are data-rich and nonhierarchical,
as long as we don't exclude ground truth. In other applications,
prior can matter a lot. This is a feature, not a bug: it's
revealing hidden structure that was implied by your modeling
assumptions and you just didn't realize it. Predictive checks to
verify you're not excluding any plausible values.
\item Marginalization/expectations.
Analytically hard, trivial with sampling.
\item MCMC.
Lot's of fun here, and deep analogies to stat mech and
Hamiltonian mechanics worth exploring. In a perfect world, you'd
wave a magic wand and generate independent draws from target
posterior. But we don't know how. Next best: generate a Markov
chain of $N$ \textit{correlated} samples, estimate an
autocorrelation time $\tau$, then you've got roughly $N/\tau$
effective samples. Obviously with some assumptions.
\item Posterior predictive checks.
So the posterior is fine, but is the model actually not wrong?
PPC is essential to verify that your model is a plausible
generative process for the data.
\item Discuss identifiability, degeneracy, and failed PPCs in main text. Better thru example anyway.
\end{itemize}
}
